---
title: "A shift in the number of surfers frequenting Oahu reef breaks before, during, and after the COVID-19 pandemic"
author: "Kyleigh Kuball"
date: "`r Sys.Date()`"
output: 
  github_document 
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(plyr)
library(dplyr)
library(plotrix)
library(RColorBrewer)
library(stringr)
library(weights)
library(ggpubr)
library(gridExtra)
library(grid)
library(gridtext)
```
 
## Abstract

With the arrival of the Covid-19 pandemic came the stay-at-home orders, and the world changed its habits. Oahu was no exception, and people were no longer at work. This study aimed to answer if more people were surfing during the 2020 stay-at-home order than in prior years. It also aimed to see if there was a greater influx of surfers in 2021 compared to earlier years. Using preliminary data from Google Earth Pro and ESRI World Imagery Wayback, surfers were counted using the ImageJ multi-point tool on multiple Oahu surf breaks in 2021, 2020, 2019, and prior years. There were few satellite images to pick from, especially with limited data in 2020, so the sample size remained small. Significance was found between the total number of surfers at Bowls and Castles (p = 0.012), Flat Island (p = 0.018), Tracks (p = 0.0076), and Waikiki (p = 0.080). There was also a significant difference between the number of surfers at Castles (p = 0.000057), Chuns (p = 0.0035), Flat Island (p = 0.000083), Kewalos (p = 0.00042), and Tracks (p = 0.000040) when compared to Waikiki. However, the small sample size led to no significance being found between any of the years. Breaks were identified with changes that might be significant if using drone imagery and further access to satellite imagery was acquired. Insignificant changes were observed in the number of people at the different breaks, and shifts were seen between the breaks. This indicates that more imagery and the possible expansion of the study to all the Hawaiian Islands might be significant. Ultimately, this study's results help identify reefs that would be beneficial to study and observations in the future regarding the influx of people and how that has affected them.

## Introduction 

Hawaiʻi is unique in how much the economy is dominated by tourism, more than anywhere in the mainland United States (Bond-Smith & Fuleky 2022). However, in March 2020, Oahu went into lockdown with a stay-at-home order. Because of this, the stay-at-home order had an enormous impact on employment since tourism was the source of most jobs in Hawaiʻi (Bond-Smith & Fuleky 2022). In addition, Hawaiʻi was slow to reopen compared to the mainland U.S., and gatherings and public events were restricted well after many states had reopened (Hallas et al. 2021). Due to the pandemic, there was likely a shift in the number of surfers utilizing the surf breaks around Hawaiʻi. This shift might be used to explore further the pandemic's impact on the vital coral reefs of the islands within Hawaiʻi. 

In 2020, people around the island of Oahu had fewer tourists around and were still cautious of the pandemic and were trying to follow many restrictions placed on activities and gatherings. Few options were available during this time, and mental health can plummet with less outdoor activity (Jackson et al., 2021). Any attempts to be outdoors on Oahu at this time were met with restrictions such as wearing masks while running and not being within 6 feet of people (like many other places). Recreation on the beaches was prohibited unless it was subsistence fishing, exercise, or water-related activities. At the start of the pandemic, there were very few who were moving outside of their homes, and local mobility, as seen from Google's phone mobility data, showed that Hawaiʻi residents were the fifth highest in staying at home when compared to the mainland states (Hallas et al. 2021). By the summer of 2020, however, people were eager to leave and, while still under restrictions, were leaving their homes (Hallas et al., 2021). Surfing likely saw a boom in the number of residents paddling out on the surf breaks during this time. 

The surf breaks began to see more locals, and places that used to be high tourism saw a shift in the number of surfers out on the surf breaks. With tourism being greatly restricted, many people unemployed, and strict restrictions on gatherings and activities, surfing was one of the few activities locals could do outside their homes (Bond-Smith & Fuleky, 2022). People were also experiencing "lockdown fatigue" and were looking to return to a new normal, as seen by the non-essential retail that began to increase during that time (Hallas et al., 2021).

Different breaks likely saw a shift in the number of people surfing in the summer of 2020. Waikiki, for instance, most likely saw a drop in the number of people since most surfing tourism tends to frequent Canoes and Queens breaks. Closer to the west of Waikiki, the surfer population likely increased at local town breaks like Bowls and Rockpiles. West side and north shore breaks probably increased since more local traffic is seen there. 

While 2020 saw an enormous decrease in the number of tourists, the following summer, there was a boom in the number of people visiting the islands. After the reduction in 2020, tourism reached a record number in the summer of 2021 by mainland U.S. visitors (Bond-Smith & Fuleky 2022). This led to yet another shift on the surf breaks, and Waikiki likely saw an uptick in the number of people, while some of the local breaks also may have seen a change upwards as people may have shifted from their surf breaks frequented during the pandemic. 

The fluxes in surfing would be a significant find since surf breaks are formed by the coral reefs estimated to provide 363.5 million per year in 2004 (Cesar & Beukering 2004). Fluxes in people around them might suggest reef growth that should be explored in the future since with a higher number of surfers can come reef damage, chemical pollutants, and more value placed on the reefs that create the popular breaks. 

This study aims to identify differences in the number of surfers at seven different surf breaks around the island of Oahu during 2020 and 2021 compared to 2019 and previous years. By identifying differences in the number of surfers, this study might open a path for future research on reefs that have seen more drastic changes in the last four years. 


## Methods

The breaks studied were all located on Oahu since preliminary observations had been made there. The breaks studied were Queens and Canoes in Waikiki, Rockpiles to Bowls at the end of Waikiki, Tracks near Nanakuli, Chuns on North Shore, Flat Island in Kailua, Castles in Kailua, and Kewalo Basin near Ala Moana. There are seven sites in total around the island of Oahu.   

Data sources included spatial imagery taken from either Google Earth Pro (7.3.4.8642) or ESRI ArcGIS World Imagery Wayback (https://livingatlas.arcgis.com/wayback/). This study sampled seven sites to compensate for the low number of images available. The chosen sites had the most open-source spatial imagery data available in less than 3m resolution while also encompassing the island of Oahu. Less west side surf breaks were available due to a lack of updated imagery (1 image per 2+ years), and more were used from the south shore since image updates occur more frequently (1-3 images a year) in those locations. In addition, due to the lack of imagery updates, only one image could be chosen per year, even at the frequently updated locations. This was due to either cloud cover, no surf, or no updates.
Images from either program were saved when the surf was evident in 2021, 2020, and 2019, and the next usable image was from years prior. Since 1-3 images were available per year and in many of them, the surf was flat/non-existent, one image was obtained per year on a day when the surf was evident, and surfers could be identified. Images were taken in different years at each break prior to 2019 due to a lack of quality imagery that both had surf and a clear view of the surfers. The years before 2019 were taken as a control to have an extra image to compare the surfer counts from 2020 and 2021. All images and counts created are available on GitHub (https://github.com/kyleighk/MBIO_630_FINAL_project) in the "counts" and "images" files.

The images were analyzed using ImageJ (1.53t) and the multi-point tool. Since surfers could easily be identified, only boards were labeled using the tool, excluding canoes, swimmers, and bodyboards. Paddleboards were likely included in the count since paddleboards are challenging to distinguish from a large surfboard in a satellite view of a lineup. All counts and images of the counts were saved in a CSV file and a jpg file to the "count" folder on GitHub. 

The CSV files were then uploaded to R studio (2022.07.1) for analysis. A table was made for each site to compare 2020 and 2021 to prior years. All the site data was then compiled, and weights were given to each image depending on if the image was taken on a weekday or a weekend. The weights assigned to weekdays were (.826) and weekends were (.325) based on the United States Bureau of Labor Statistics for 2021, in which the total adults 21 and up that worked weekdays tended to average around 82.6%. The total number of adults 21 and up that worked weekend days was approximately 32.5% (U.S. Bureau of Labor Statistics, 2022). A t-test and a weighted t-test were conducted between 2020 and prior years, 2021 and prior years, and between 2020 and 2021. 


## Results

The data from ImageJ was compiled by site with a data table showing the surf break name,  the date the image was taken, the year the image was taken, the count of surfers, and the day the image was taken. This information was then compiled into Figure 1, showing the four time points for each of the seven sites. Figure 1 shows the number of surfers counted at each site at each time point and is shaded with the day of the week the image was taken. Figure 1 also highlights the time of the week that images were available. 

An ANOVA and Tukey post hoc analysis were conducted to identify differences between the sites and the years from the data shown in Figure 1. Significant differences were found between the count of surfers between Castles (p = 0.012), Flat Island (p = 0.018), Tracks (p = 0.0076), and Waikiki (p = 0.080) when compared to Bowls. Significant differences were also found when Castles (p = 0.000057), Chuns (p = 0.0035), Flat Island (p = 0.000083), Kewalos (p = 0.00042), and Tracks (p = 0.000040) were compared to Waikiki. When comparing the difference in the surfers counted year to year at each site, no significant differences were found. 

Figure 2 shows the weighted mean number of surfers found each year with the number of images that were used to obtain this mean. Since only a small number of images were available, both a t-test and weighted t-test were used to compare the total counts from 2020 and 2021 to the prior years with the most counts, 2019, 2018, and 2014. No significant differences were found between 2020 and 2019 (t = -0.81, p = 0.44; weighted t = -0.80, p = 0.45), 2018 (t = 0.74, p = 0.48; weighted t = 0.66, p = 0.53), and 2014 (t = -0.18, p = 0.86). Similar insignificance was found between 2021 and 2019 (t = -0.93, p = 0.38, weighted t = -1.00, p = 0.34), 2018 (t = 0.53, p = 0.61, weighted t = 0.39, p = 0.71), and 2014(t = -0.29, p = 0.78, weighted t = -0.81, p = 0.44). The count differences were also compared between 2020 and 2021(t = -0.16, p = 0.87, weighted t = -0.30, p = 0.77) and no significance was found. 

Figure 3 illustrates the shifts in the total counts over the years that the images were taken. The bottom illustration shows the proportions of the total number of surfers found before, during, and after the stay-at-home order at the different surf sites. While the significance is below statistical power, slight growth in 2021 can be seen in Waikiki and Castles. Meanwhile, Bowls appears to have the greatest number of surfers during 2020, and Chuns and Tracks appear to have kept a steady number of surfers except for 2021. In addition, Flat Island appears to have had a pulse of surfers in 2014. 


```{r analysis chunk 1, echo=FALSE}
waikiki_8_12_2014 <- read.csv("counts/Waikiki_8_12_2014.csv") #Tuesday
waikiki_6_10_2019 <- read.csv("counts/Waikiki_6_10_2019.csv") #Monday
waikiki_9_28_2020 <- read.csv("counts/Waikiki_ESRI_9_28_2020.csv") #Monday
waikiki_6_2_2021 <- read.csv("counts/Waikiki_6_2_2021.csv") #Wednesday

name1 <- c("waikiki", "waikiki", "waikiki", "waikiki")
date1 <- c("8/12/2014", "6/10/2019", "9/28/2020", "6/2/2021")
year1 <- c("2014", "2019", "2020", "2021")
day1 <- c("Tuesday", "Monday", "Monday", "Wednesday")
count1 <- c(length(waikiki_8_12_2014$X.1), length(waikiki_6_10_2019$X.1), length(waikiki_9_28_2020$X.1), length(waikiki_6_2_2021$X.1) )

waikiki <- data.frame(name = name1, date = date1, year = year1, count = count1, day = day1)

std_error1 <- std.error(waikiki$count)

waikiki_plot <- ggplot(data = waikiki, aes(x = date, y = count, fill = day, label = count)) +
  geom_bar(stat = "identity") +
  geom_text(size = 10, position = position_stack(vjust = 0.25)) +
  geom_errorbar(aes(date, ymin = count - std_error1, ymax = count + std_error1)) +
  scale_fill_brewer(palette = 3, type = "qual") +
  scale_x_discrete(limits = c("8/12/2014", "6/10/2019", "9/28/2020", "6/2/2021")) +
  ggtitle("Waikiki") +
  theme(plot.caption = element_text(hjust = 0.5), plot.title = element_text(hjust = 0.5), text = element_text(size = 25))
```

```{r, echo=FALSE}
tracks_12_20_2015 <- read.csv("counts/Tracks_ESRI_12_20_2015.csv") #Sunday
tracks_2_18_2018 <- read.csv("counts/Tracks_2_18_2018.csv") #Sunday
tracks_10_3_2020 <- read.csv("counts/Tracks_ESRI_10_3_2020.csv") #Saturday
tracks_3_18_2021 <- read.csv("counts/Tracks_3_18_2021.csv") #Thursday

name2 <- c("tracks", "tracks","tracks", "tracks")
date2 <- c("12/20/2015", "2/18/2018","10/3/2020", "3/18/2021")
year2 <- c("2015", "2018","2020", "2021")
day2 <- c("Sunday", "Sunday", "Saturday", "Thursday")
count2 <- c(length(tracks_12_20_2015$X.1), length(tracks_2_18_2018$X.1), length(tracks_10_3_2020$X.1), length(tracks_3_18_2021$X.1))

tracks <- data.frame(name = name2, date = date2, year = year2, count = count2, day = day2)

std_error2 <- std.error(tracks$count)

tracks_plot <- ggplot(data = tracks, aes(x = date, y = count, fill = day, label = count)) +
  geom_bar(stat = "identity") +
  geom_text(size = 10, position = position_stack(vjust = 0.25)) +
  geom_errorbar(aes(date, ymin = count - std_error2, ymax = count + std_error2)) +
  scale_fill_brewer(palette = 3, type = "qual") +
  scale_x_discrete(limits = c("12/20/2015", "2/18/2018","10/3/2020", "3/18/2021")) +
  ggtitle("Tracks") +
  theme(plot.caption = element_text(hjust = 0.5), plot.title = element_text(hjust = 0.5), text = element_text(size = 25))
```

```{r, echo=FALSE}
kewalos_1_14_2018 <- read.csv("counts/Kewalos_1_14_2018.csv") #Sunday
kewalos_5_26_2019 <- read.csv("counts/Kewalos_5_26_2019.csv") #Sunday
kewalos_1_22_2020 <- read.csv("counts/Kewalos_1_22_2020.csv") #Wednesday
kewalos_6_2_2021 <- read.csv("counts/Kewalos_6_2_2021.csv") #Wednesday

name3 <- c("kewalos", "kewalos","kewalos", "kewalos")
date3 <- c("1/14/2018", "5/26/2019","1/22/2020", "6/2/2021")
year3 <- c("2018", "2019","2020", "2021")
day3 <- c("Sunday", "Sunday", "Wednesday", "Wednesday")
count3 <- c(length(kewalos_1_14_2018$X.1), length(kewalos_5_26_2019$X.1), length(kewalos_1_22_2020$X.1), length(kewalos_6_2_2021$X.1))

kewalos <- data.frame(name = name3, date = date3, year = year3, count = count3, day = day3)

std_error3 <- std.error(kewalos$count)

kewalos_plot <- ggplot(data = kewalos, aes(x = date, y = count, fill = day, label = count)) +
  geom_bar(stat = "identity") +
  geom_text(size = 10, position = position_stack(vjust = 0.25)) +
  geom_errorbar(aes(date, ymin = count - std_error3, ymax = count + std_error3)) +
  scale_fill_brewer(palette = 3, type = "qual") +
  scale_x_discrete(limits = c("1/14/2018", "5/26/2019","1/22/2020", "6/2/2021")) +
  ggtitle("Kewalos") +
  theme(plot.caption = element_text(hjust = 0.5), plot.title = element_text(hjust = 0.5), text = element_text(size = 25))
```

```{r, echo=FALSE}
flatisland_7_17_2006 <- read.csv("counts/Flatisland_ESRI_7_17_2006.csv") #Monday
flatisland_8_10_2014 <- read.csv("counts/Flatisland_8_10_2014.csv") #Sunday
flatisland_9_29_2020 <- read.csv("counts/Flatisland_ESRI_9_29_2020.csv")#Tuesday 
flatisland_3_25_2021 <- read.csv("counts/Flatisland_3_25_2021.csv") #Thursday

name4 <- c("flatisland", "flatisland","flatisland", "flatisland")
date4 <- c("7/17/2006", "8/10/2014","9/29/2020", "3/25/2021")
year4 <- c("2006", "2014","2020", "2021")
day4 <- c("Monday", "Sunday", "Tuesday", "Thursday")
count4 <- c(length(flatisland_7_17_2006$X.1), length(flatisland_8_10_2014$X.1), length(flatisland_9_29_2020$X.1), length(flatisland_3_25_2021$X.1))

flatisland <- data.frame(name = name4, date = date4, year = year4, count = count4, day = day4)

std_error4 <- std.error(flatisland$count)

flatisland_plot <- ggplot(data = flatisland, aes(x = date, y = count, fill = day, label = count)) +
  geom_bar(stat = "identity") +
  geom_text(size = 10, position = position_stack(vjust = 0.25)) +
  geom_errorbar(aes(date, ymin = count - std_error4, ymax = count + std_error4)) +
  scale_fill_brewer(palette = 3, type = "qual") +
  scale_x_discrete(limits = c("7/17/2006", "8/10/2014","9/29/2020", "3/25/2021")) +
  ggtitle("Flat Island") +
  theme(plot.caption = element_text(hjust = 0.5), plot.title = element_text(hjust = 0.5), text = element_text(size = 25))
```


```{r, echo=FALSE}
chuns_8_26_2018 <- read.csv("counts/Chuns_ESRI_8_26_2018.csv") #Sunday
chuns_5_26_2019 <- read.csv("counts/Chuns_5_26_2019.csv") #Sunday
chuns_8_3_2020 <- read.csv("counts/Chuns_ESRI_8_3_2020.csv") #Friday
chuns_3_18_2021 <- read.csv("counts/Chuns_3_18_2021.csv") #Thursday

name5 <- c("chuns", "chuns","chuns", "chuns")
date5 <- c("8/26/2018", "5/26/2019","8/3/2020", "3/18/2021")
year5 <- c("2018", "2019","2020", "2021")
day5 <- c("Sunday", "Sunday", "Friday", "Thursday")
count5 <- c(length(chuns_8_26_2018$X.1), length(chuns_5_26_2019$X.1), length(chuns_8_3_2020$X.1), length(chuns_3_18_2021$X.1))

chuns <- data.frame(name = name5, date = date5, year = year5, count = count5, day = day5)

std_error5 <- std.error(chuns$count)

chuns_plot <- ggplot(data = chuns, aes(x = date, y = count, fill = day, label = count)) +
  geom_bar(stat = "identity") +
  geom_text(size = 10, position = position_stack(vjust = 0.25)) +
  geom_errorbar(aes(date, ymin = count - std_error5, ymax = count + std_error5)) +
  scale_fill_brewer(palette = 3, type = "qual") +
  scale_x_discrete(limits = c("8/26/2018", "5/26/2019","8/3/2020", "3/18/2021")) +
  ggtitle("Chuns") +
  theme(plot.caption = element_text(hjust = 0.5), plot.title = element_text(hjust = 0.5), text = element_text(size = 25))
```

```{r, echo=FALSE}
castles_1_29_2013 <- read.csv("counts/castles_1_29_2013.csv") #Tuesday
castles_8_10_2014 <- read.csv("counts/Castles_8_10_2014.csv") #Sunday
castles_6_2020 <- read.csv("counts/Castles_6_2020.csv") #?Monday
castles_3_25_2021 <- read.csv("counts/Castles_3_25_2021.csv") #Thursday

name6 <- c("castles", "castles","castles", "castles")
date6 <- c("1/29/2013", "8/10/2014","6/1/2020", "3/25/2021")
year6 <- c("2013", "2014","2020", "2021")
day6 <- c("Tuesday", "Sunday", "Monday", "Thursday")
count6 <- c(length(castles_1_29_2013$X.1), length(castles_8_10_2014$X.1), length(castles_6_2020$X.1), length(castles_3_25_2021$X.1))

castles <- data.frame(name = name6, date = date6, year = year6, count = count6, day = day6)

std_error6 <- std.error(castles$count)

castles_plot <- ggplot(data = castles, aes(x = date, y = count, fill = day, label = count)) +
  geom_bar(stat = "identity") +
  geom_text(size = 10, position = position_stack(vjust = 0.25)) +
  geom_errorbar(aes(date, ymin = count - std_error6, ymax = count + std_error6)) +
  scale_fill_brewer(palette = 3, type = "qual") +
  scale_x_discrete(limits = c("1/29/2013", "8/10/2014","6/1/2020", "3/25/2021")) +
  ggtitle("Castles") +
  theme(plot.caption = element_text(hjust = 0.5), plot.title = element_text(hjust = 0.5), text = element_text(size = 25))
```


```{r, echo=FALSE}
bowls_1_14_2018 <- read.csv("counts/Bowls_1_14_2018.csv") #Sunday
bowls_6_10_2019 <- read.csv("counts/Bowls_6_10_2019.csv") #Monday
bowls_9_28_2020 <- read.csv("counts/Bowls_ESRI_9_28_2020.csv")#Monday
bowls_6_2_2021 <- read.csv("counts/Bowls_6_2_2021.csv")#Wednesday

name7 <- c("bowls", "bowls", "bowls", "bowls")
date7 <- c("1/14/2018", "6/10/2019","9/28/2020", "6/2/2021")
year7 <- c("2018", "2019", "2020", "2021")
day7 <- c("Sunday", "Monday", "Monday", "Wednesday")
count7 <- c(length(bowls_1_14_2018$X.1), length(bowls_6_10_2019$X.1), length(bowls_9_28_2020$X.1), mean(length(bowls_6_2_2021$X.1)))

bowls <- data.frame(name = name7, date = date7, year = year7, count = count7, day = day7)

std_error7 <- std.error(bowls$count)

bowls_plot <- ggplot(data = bowls, aes(x = date, y = count, fill = day, label = count)) +
  geom_bar(stat = "identity") +
  geom_text(size = 10, position = position_stack(vjust = 0.25)) +
  geom_errorbar(aes(date, ymin = count - std_error7, ymax = count + std_error7)) +
  scale_fill_brewer(palette = 3, type = "qual") +
  scale_x_discrete(limits = c("1/14/2018", "6/10/2019","9/28/2020", "6/2/2021")) +
  ggtitle("Bowls") +
  theme(plot.caption = element_text(hjust = 0.5), plot.title = element_text(hjust = 0.5), text = element_text(size = 25))
```

```{r, echo=FALSE, fig.width=27, fig.height= 18}

t <- textbox_grob("**Figure 1.** Total counts of surfers at each of the study sites. Dates of images taken are shown on the x-axis and total count of surfers is shown on the y-axis. Colors of the charts coordinate with days of the week and are shown in the legend.", gp = gpar(fontsize = 30))

grid.arrange(waikiki_plot, flatisland_plot, bowls_plot, castles_plot, chuns_plot, kewalos_plot, tracks_plot, bottom = t, nrow=3)
```




<br />
<br />
<br />

```{r, echo=FALSE}
surf.dat <- rbind(waikiki, flatisland, bowls, castles, chuns, kewalos, tracks)

surf.dat$weight <- ifelse(surf.dat$day == 'Saturday', .325,
                          ifelse(surf.dat$day == 'Sunday', .325, .826))

surf.dat$year <- as.character(surf.dat$year)
```



```{r, echo=FALSE, include=FALSE}
TukeyHSD(aov(count ~ name + year, data = surf.dat))
```

```{r, echo=FALSE, warning=FALSE, fig.width=10, fig.height= 4}
weighted_average <- ddply(surf.dat, .(year), function(x) data.frame(w_mean_surfers = weighted.mean(x$count, x$weight)))

total_per_year <- count(surf.dat$year)

summary <- data.frame(weighted_average, num_observ = total_per_year$freq)

figure_2 <- ggplot(data = summary, aes(x = year, y = w_mean_surfers, fill = num_observ)) +
  geom_col(postion = "fill")

t2 <- textbox_grob("**Figure 2.** This figure illustrates the number of observations collected for each year. The year is located on the x-axis and the weighted mean number of surfers is on the y-axis. The weighted mean was calculated using the United States Bureau of Labor Statistics for 2021 by assigning a weight of 0.826 for weekdays and 0.325 for weekends (U.S. Bureau of Labor Statistics 2022). The t-tests conducted were based on 2019, 2018, and 2014 having the greatest number of open source imagery prior to 2020.")

grid.arrange(figure_2, t2, nrow=2)
```

<br />
<br />
<br />

```{r, echo=FALSE, include=FALSE}
df <- subset(surf.dat, year %in% c("2020","2021"))

coefficients(summary(lm(count ~ year,data = df)))

t.test(df$count[df$year =="2020"], df$count[df$year =="2021"],data = df)

coefficients(summary(lm(count ~ year, data = df, weight = df$weight)))
```

```{r, echo=FALSE, include=FALSE}
df2 <- subset(surf.dat, year %in% c("2019","2020")) 

coefficients(summary(lm(count ~ year,data = df2)))

t.test(df2$count[df2$year =="2019"], df2$count[df2$year =="2020"], data = df2)

coefficients(summary(lm(count ~ year, data = df2, weight = df2$weight)))
```

```{r, echo=FALSE, include=FALSE}
df3 <- subset(surf.dat, year %in% c("2018","2020")) 

coefficients(summary(lm(count ~ year,data = df3)))

t.test(df3$count[df3$year =="2018"], df3$count[df3$year =="2020"], data = df3)

coefficients(summary(lm(count ~ year, data = df3, weight = df3$weight)))
```

```{r, echo=FALSE, include=FALSE}
df4 <- subset(surf.dat, year %in% c("2014","2020")) 

coefficients(summary(lm(count ~ year,data = df4)))

t.test(df4$count[df4$year =="2014"], df4$count[df4$year =="2020"], data = df4)

coefficients(summary(lm(count ~ year, data = df4, weight = df4$weight)))
```

```{r, echo=FALSE, include=FALSE}
df <- subset(surf.dat, year %in% c("2019","2021"))

coefficients(summary(lm(count ~ year,data = df)))

t.test(df$count[df$year =="2019"], df$count[df$year =="2021"],data = df)

coefficients(summary(lm(count ~ year, data = df, weight = df$weight)))
```

```{r, echo=FALSE, include=FALSE}
df <- subset(surf.dat, year %in% c("2014","2021"))

coefficients(summary(lm(count ~ year,data = df)))

t.test(df$count[df$year =="2014"], df$count[df$year =="2021"],data = df)

coefficients(summary(lm(count ~ year, data = df, weight = df$weight)))
```

```{r, echo=FALSE,include=FALSE }
df <- subset(surf.dat, year %in% c("2018","2021"))

coefficients(summary(lm(count ~ year,data = df)))

t.test(df$count[df$year =="2018"], df$count[df$year =="2021"],data = df)

coefficients(summary(lm(count ~ year, data = df, weight = df$weight)))
```





```{r, echo=FALSE, warning=FALSE, fig.width=10, fig.height=7}
pandemic <- filter(surf.dat, year > 2018)
plot1 <- ggplot(data = surf.dat, aes(x = name, y = count, fill = year)) +
  geom_col(postion = "fill")

plot2 <- ggplot(data = pandemic, aes(x = year, y = count, fill = name)) +
  geom_bar(stat = "identity") +
  scale_fill_brewer(palette = 3, type = "qual") +
  theme()

t3 <- textbox_grob("**Figure 3.** Both of these illustrations show the shift in the total counts over the course of a few years. The first graph shows how the different sites' counts changed each year. The surf sites are located on the x-axis and the total counts of surfers is located on the y-axis. The year corresponds to the color of each bar. The second illustration focuses on the pandemic years and shows a shift in the locations surfed as the pandmeic continued from 2019. The year is located on the x-axis and the count is located on the y-axis. The color of the bars correspond to the surf sites.")

grid.arrange(arrangeGrob(plot1, plot2), t3, nrow = 1)


```


## Discussion

Overall, the goal of this study was to obtain a preliminary understanding if there was a difference in the number of surfers before, during, and after the stay-at-home order on Oahu in 2020. While there was limited imagery, only 1-3 images per year, a beginning summary was obtained. There appeared to be a shift in surfers around the breaks of Oahu, and the imagery suggests that with further data, a statistical significance might be found in the number of surfers per year between 2019, 2020, and 2021. 

As expected, the number of people between the breaks was statistically significant, with Bowls and Waikiki varying in numbers from the other five breaks (Figure 1). The surfer population tends to run high at these breaks, so when compared to breaks like Tracks, Flat Island, or Castles, it seems logical that the difference is statistically significant. The days of the week also seemed to influence the more popular breaks, with Chuns having a greater number of people on Sunday and Friday than on Thursday (Figure 1). Tracks, Kewalos, and Flat Island appear to have a shift in people from weekdays to weekend days as well. Waikiki does not seem to be as impacted by the day of the week, with weekdays having 100+ people. Similarly, Bowls had the most people counted on Mondays. 

The shift from year to year is hard to determine with so few images; however, this data does suggest further investigations. The Tukey post hoc test used did not determine any significance between the years. However, the graphs (Figures 1 & 3) suggest there might have been a growth in the total number of surfers during 2021 and a decline in 2020 at Waikiki and Castles, while other sites such as Bowls may have seen growth in 2020 and a decline in 2021. Waikiki and Castles appear to have had a greater number of people in 2021 than in 2020 or 2019. The lowest number of surfers for Waikiki was unexpectedly in 2019 prior to the stay-at-home order and temporary loss of tourism. Considering that there was a loss of tourism, surf schools, and board rentals, with further image counts, this number might change. Kewalos appears to have a decline in 2020, which is expected since the surf schools and board rentals were not operating at that time. The higher count of surfers at Bowls in 2020 might be attributed to it being a very popular spot for town locals and, with the loss of jobs, an easy break to get to. 

The lack of imagery data was surprising. Given that many popular inland locations are updated relatively frequently, it was assumed that this would be similar to locations such as Waikiki. Yet, even as popular as Waikiki is, Google Earth Pro had some years in which the same image was used in a 1–2-year streak. The same was seen in the ESRI ArcGIS imagery. The hope of the study was to compare multiple images within a year. However, that had to be adjusted. In the future, a broader range will be used, such as the main Hawaiian Islands rather than just Oahu, to increase the site sample size. The years studied will be adjusted as well so that more imagery can be found at each site both on weekdays and weekends, and more data will be available to compare to 2020 and 2021. Finally, historic drone imagery (such as that taken by surf photographers) or more frequent sensing mechanisms will be explored to get a better picture of how the populations at the surf breaks are changing. 

In all, this study laid the groundwork for further research into the human populations frequenting Hawaiian surf breaks, which in turn will provide a reference when studying changes to the coral reefs that make up the surf breaks. It is the hope of this study that this will contribute to the research of human impact on corals and what can be done to protect the reefs that do so much.


## References

Bond-Smith, S., & Fuleky, P. (2022). The pandemic's effects on Hawaii's economy (No. 2022-4).

Cesar, H. S. J. & van Beukering, P. J. H. Economic valuation of the coral reefs of Hawai'i. Pacific Sci. 58, 231–242 (2004).

Hallas, L., Hatibie, A., Koch, R., Majumdar, R., Pyarali, M., Wood, A., Hale, T. (2021) "Variation in US states' responses to COVID-19", BSG-WP-2020/034 Version 3.0, Blavatnik School of Government Working Paper Series, University of Oxford.

Jackson, S. B., Stevenson, K. T., Larson, L. R., Peterson, M. N., & Seekamp, E. (2021). Outdoor activity participation improves adolescents' mental health and well-being during the COVID-19 pandemic. International Journal of Environmental Research and Public Health, 18(5), 2506.

U.S. Bureau of Labor Statistics. (2022, June 23). American Time use Survey Summary. Retrieved October 7 2022, from https://www.bls.gov/charts/american-time-use/emp-by-ftpt-job-edu-p.htm




